---
title: "Heapsort"
author: "xhan09"
output:
  pdf_document: 
    latex_engine: xelatex
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6, fig.height = 4, fig.align = 'center')
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

## Problem 2: Implementing priority queues and sorting

###2.1 The index of the parent node of element i is (i/2). The indices of the left and right child nodes if element i are 2i and (2i+1), respectively.

### 2.2 Following is the function make_heap to return an empty heap with length LMAX.

```{r}
make_heap <- function(LMAX) {
  heap <- rep(0, LMAX)
  return(heap)
}
```

### 2.3 Following is the function max_heap to return the maximum element of the heap.

```{r}
max_heap <- function(heap) {
  return(heap[1])
}
max_heap(c(9,8,7,4,1,1,3,2,2))
```

### 2.4 Following is the function remove_max to remove the maximum element of the heap.

```{r}
remove_max <- function(heap) {
  len <- length(heap)
  max_val <- heap[1]
  last_val <- heap[len]
  newheap <- heap[-len]
  newheap[1] <- last_val
  
  i <- 1
  while (2*i <= length(newheap)) {
    left_child <- newheap[2*i]
    right_child <- newheap[2*i + 1]
    child_list <- c(left_child, right_child)
    child_list <- child_list[!is.na(child_list)]
    
    if (newheap[i] >= max(child_list)) {break}
    
    if (is.na(right_child) || left_child >= right_child) {
      newheap[2*i] <- newheap[i]
      newheap[i] <- left_child
      i <- 2*i
    }
    else {
      newheap[2*i + 1] <- newheap[i]
      newheap[i] <- right_child
      i <- 2*i + 1
    }
  }
  return(list(V=max_val, H=newheap))
}
remove_max(c(9,8,7,4,1,1,3,2,2))
```

### 2.4 Following is the function insert_element to insert a new element into the heap.

```{r}
insert_element <- function(heap, new_element) {
  if (identical(which(heap != 0), integer(0))) {newindex <- 1}
  else {newindex <- max(which(heap != 0)) + 1}
  heap[newindex] <- new_element
  
  i <- newindex
  while (floor(i/2) >= 1) {
    parent_node = floor(i/2)
    if (heap[i] > heap[parent_node]) {
      tmp <- heap[parent_node]
      heap[parent_node] <- heap[i]
      heap[i] <- tmp
      i <- parent_node
    } else {break}
  }
  return(heap)
}
insert_element(c(8,4,7,2,1,1,3,2,0), 9)
```

### 2.5 Following is the function heap_sort to sort a vector of numbers using the functions defined above.

```{r}
heap_sort <- function(input) {
  LMAX <- length(input)
  prevheap <- make_heap(LMAX)
  
  for (i in 1:LMAX) {
    curheap <- insert_element(prevheap, input[i])
    prevheap <- curheap
  }

  newvector <- rep(0, LMAX)
  for (i in 1:LMAX) {
    tmp <- remove_max(prevheap)
    newvector[i] <- tmp$V
    prevheap <- tmp$H
  }
  return(newvector)
}

inheap <- sample(1:100, 20)
inheap
heap_sort(inheap)
```
