---
title: "Power Method"
author: "xhan09"
output:
  pdf_document: 
    latex_engine: xelatex
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6, fig.height = 4, fig.align = 'center')
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

## Problem 1: The power method

### 1.1 I have generated a random symmetric 5 x 5 matrix from a uniform distribution. Eigenvalue and eigenvectors for this matrix are shown below

```{r}
A <- matrix(runif(25), 5, 5)
A <- A %*% t(A)
eigen(A)
```

### 1.2 I have written the function "pmethod"" to return the first eigenvector and eigenvalue of the input matrix. We can see the result from this function is the same as the result from eigen(A).

```{r}
pmethod <- function(imatrix) {
  uold <- c(1, 1, 1, 1, 1)
  eps=1e-6
  step <- 1
  maxstep <- 1000
  repeat {
    unew <- imatrix %*% uold
    unew <- unew/sqrt(sum(unew^2))
    if(norm(abs(unew)-abs(uold)) <= eps) break
    uold <- unew
    if (step == maxstep) break
    step <- step + 1
  }
  eigenvalue <- (t(unew) %*% imatrix %*% unew) / (t(unew) %*% unew)
  return(list(vector = unew, value = eigenvalue[1,1]))
}
result <- pmethod(A)
result
```

### 1.3 To obtain the second largest eigenvalue and eigenvector, we have done a transformation to A to generate a matrix "tmatrix", whose first eigenvalue is 0. In this case, the largest eigenvalue for "tmatrix" corresponds to the second largest eigenvalue of A.

```{r}
tmatrix <- A - result$value * result$vector %*% t(result$vector)
tmatrix
result2 <- pmethod(tmatrix)
result2
```